---
title: "Calculating Clinical Scores"
author: "clifR Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating Clinical Scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r}
library(clifR)
library(dplyr)
library(ggplot2)
```

# Overview

clifR reproduces three widely used severity and comorbidity indices from clifpy:

- **SOFA** (Sequential Organ Failure Assessment) — six organ-system component scores
  plus a total, computed from a wide, time-aligned dataset.
- **Charlson Comorbidity Index (CCI)** — from ICD-10-CM hospital-diagnosis codes.
- **Elixhauser Comorbidity Index** — from the same diagnosis codes, using the
  Quan (2011) mapping with van Walraven weights.

Integer-valued scores match clifpy exactly; there is no tolerance budget for them.

# SOFA scores

## The medication-unit prerequisite

The cardiovascular component of SOFA depends on vasopressor infusion rates expressed
in `mcg/kg/min`. The wide-dataset builder therefore expects the continuous-medication
doses to have been **standardized to preferred units first**, so that vasopressor
columns arrive named `norepinephrine_mcg_kg_min`, `epinephrine_mcg_kg_min`, and so
on. If you skip this step, SOFA cannot find those columns.

We convert the four SOFA vasopressors to `mcg/kg/min`. `override = TRUE` lets the
converter warn (rather than stop) on medication categories it cannot express in an
acceptable target unit, and `save_to_table = FALSE` returns the converted frame,
which we place back on the medication table so the wide builder picks it up.

```{r}
clif <- load_demo_clif()

conversion <- clif$convert_dose_units_for_continuous_meds(
  preferred_units = list(
    norepinephrine = "mcg/kg/min",
    epinephrine    = "mcg/kg/min",
    dopamine       = "mcg/kg/min",
    dobutamine     = "mcg/kg/min"
  ),
  override = TRUE,
  save_to_table = FALSE
)

# Attach the converted infusion rates so the wide builder emits *_mcg_kg_min columns.
clif$medication_admin_continuous$df <- conversion$converted
```

## Computing SOFA

`compute_sofa_scores()` builds the wide dataset it needs (vitals, labs, respiratory
support, GCS assessments, and the converted vasopressors), takes the worst value per
patient, and returns one row of component scores per grouping ID. Here we group by
`hospitalization_id`.

```{r}
sofa_scores <- clif$compute_sofa_scores(id_name = "hospitalization_id")

sofa_scores |>
  select(hospitalization_id, starts_with("sofa_")) |>
  head()
```

The component columns are the six organ systems plus the total:

```{r}
sofa_scores |>
  select(sofa_cv_97, sofa_coag, sofa_liver, sofa_resp, sofa_cns, sofa_renal, sofa_total) |>
  summary()
```

```{r fig.alt="Histogram of total SOFA scores across hospitalizations"}
ggplot(sofa_scores, aes(x = sofa_total)) +
  geom_histogram(binwidth = 1, fill = "#009E73", colour = "white") +
  labs(
    title = "Distribution of total SOFA scores",
    x = "Total SOFA",
    y = "Hospitalizations"
  ) +
  theme_minimal()
```

## Grouping by encounter block

Passing `id_name = "encounter_block"` (the default) scores stitched encounters rather
than individual hospitalizations; clifR runs encounter stitching automatically when
the wide dataset lacks an `encounter_block` column.

```{r eval=FALSE}
sofa_by_encounter <- clif$compute_sofa_scores(id_name = "encounter_block")
```

# Comorbidity indices (CCI and Elixhauser)

Both indices are computed from the `hospital_diagnosis` table, which is **not part of
the bundled demo dataset**. In a real workflow you would load it and call the
orchestrator helpers:

```{r eval=FALSE}
clif$load_table("hospital_diagnosis")
cci  <- clif$compute_cci_scores()   # calculate_cci() under the hood
elix <- clif$compute_elix_scores()  # calculate_elix() under the hood
```

The underlying free functions, `calculate_cci()` and `calculate_elix()`, accept either
a `HospitalDiagnosis` table object or a plain data frame with `hospitalization_id`,
`diagnosis_code`, and `diagnosis_code_format` columns. To keep this vignette runnable,
we build a small illustrative diagnosis frame with real ICD-10-CM codes. (This stands
in for a loaded `hospital_diagnosis` table; the calculation itself is the real one.)

```{r}
hospital_diagnosis <- tibble::tribble(
  ~hospitalization_id, ~diagnosis_code, ~diagnosis_code_format,
  "H1", "I21.9",  "icd10cm",  # acute myocardial infarction
  "H1", "E11.9",  "icd10cm",  # type 2 diabetes
  "H1", "N18.5",  "icd10cm",  # chronic kidney disease, stage 5
  "H2", "J44.9",  "icd10cm",  # COPD
  "H2", "I50.9",  "icd10cm",  # congestive heart failure
  "H3", "C34.90", "icd10cm"   # lung cancer
)
```

## Charlson Comorbidity Index

`calculate_cci()` returns one binary flag column per Charlson condition plus the
weighted `cci_score`.

```{r}
cci <- calculate_cci(hospital_diagnosis)
cci |>
  select(hospitalization_id, cci_score)
```

```{r}
# Which conditions were detected for hospitalization H1?
cci |>
  filter(hospitalization_id == "H1") |>
  select(-hospitalization_id, -cci_score) |>
  tidyr::pivot_longer(everything(), names_to = "condition", values_to = "present") |>
  filter(present == 1)
```

## Elixhauser Comorbidity Index

`calculate_elix()` follows the same shape, returning condition flags and the weighted
`elix_score`.

```{r}
elix <- calculate_elix(hospital_diagnosis)
elix |>
  select(hospitalization_id, elix_score)
```

Both functions apply the "assign0" comorbidity hierarchy by default (for example, a
metastatic tumour suppresses the milder solid-tumour flag). Pass `hierarchy = FALSE`
to disable it.

# Session info

```{r}
sessionInfo()
```
