---
title: "Data Quality and Validation"
author: "clifR Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Quality and Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r}
library(clifR)
library(dplyr)
```

# Overview

Data quality is central to multi-site ICU research: results are only comparable if
every site's data conforms to the same CLIF specification. clifR ports clifpy's data
quality assessment (DQA) engine, which checks each table across four families of
rules:

- **Conformance** — required columns, data types, datetime formats, permissible
  categorical values, category-to-group mappings.
- **Completeness** — missingness, conditionally required fields, mCIDE value coverage.
- **Relational integrity** — foreign-key relationships across tables.
- **Plausibility** — chronological ordering, numeric-range plausibility, field
  plausibility, temporal consistency, duplicate composite keys.

This vignette works against the bundled CLIF demo data, which is CLIF 2.1. clifR also
supports CLIF 3.0, and includes a migration path from 2.1 to 3.0.

# Per-table validation

The quickest check is each table object's own `validate()` method, which validates
against its schema and records findings in `$errors`. `isvalid()` reports whether the
last run was clean.

```{r}
labs_table <- load_demo_labs()
labs_table$validate(verbose = TRUE)
labs_table$isvalid()
```

```{r}
# Inspect the first few findings, if any
length(labs_table$errors)
if (length(labs_table$errors) > 0) labs_table$errors[[1]]
```

Across an orchestrator, `validate_all()` runs this for every loaded table:

```{r}
clif <- load_demo_clif(tables = c("patient", "hospitalization", "adt", "vitals", "labs"))
invisible(clif$validate_all(verbose = TRUE))
```

# The full DQA engine

`run_full_dqa()` runs the complete DuckDB-backed check suite on a single table's data
frame and schema, returning a nested result grouped by check family. Each individual
check is a named list with `check_type`, `passed`, and lists of `errors`, `warnings`,
and `info` messages.

```{r}
dqa_result <- run_full_dqa(
  df = labs_table$df,
  schema = labs_table$schema,
  table_name = "labs",
  clif_version = "2.1"
)

names(dqa_result)
names(dqa_result$conformance)
```

## Summarizing the results

It is convenient to flatten the nested result into one row per check:

```{r}
check_families <- c("conformance", "completeness", "relational", "plausibility")

summarize_family <- function(family) {
  checks <- dqa_result[[family]]
  if (length(checks) == 0) return(NULL)
  rows <- lapply(checks, function(single_check) {
    tibble(
      family = family,
      check = single_check$check_type,
      passed = isTRUE(single_check$passed),
      n_errors = length(single_check$errors),
      n_warnings = length(single_check$warnings)
    )
  })
  bind_rows(rows)
}

dqa_summary <- bind_rows(lapply(check_families, summarize_family))
dqa_summary
```

```{r}
# Headline: how many checks passed?
dqa_summary |>
  count(passed)
```

```{r}
# Any check that raised errors or warnings
dqa_summary |>
  filter(n_errors > 0 | n_warnings > 0)
```

A single check can be inspected in full to see the underlying messages:

```{r}
str(dqa_result$plausibility$numeric_range_plausibility, max.level = 2, list.len = 4)
```

# Generating validation reports

For sharing with a site, clifR renders human-readable reports from validation output.
`generate_text_report()` writes a plain-text report for one table, and
`generate_combined_report()` assembles per-table JSON results (written to a directory)
into a combined HTML/PDF report. These write files, so they are shown but not executed
here.

```{r eval=FALSE}
# Text report for a single table
generate_text_report(
  validation_data = dqa_result,
  table_name = "labs",
  output_path = "labs_validation_report.txt",
  site_name = "Example Health System"
)

# Combined report across tables (reads per-table JSON from json_dir)
generate_combined_report(
  json_dir = "validation_json",
  output_dir = "validation_reports",
  table_names = c("patient", "hospitalization", "labs", "vitals"),
  site_name = "Example Health System"
)
```

# Adult Sepsis Event (ASE)

`compute_ase()` implements the CDC Adult Sepsis Event surveillance definition as a
DuckDB pipeline ported from clifpy. It reads directly from a CLIF data directory
(cultures, labs, medications, vitals, and more), so it needs a full dataset rather
than a single table and is shown here for reference:

```{r eval=FALSE}
ase_results <- compute_ase(
  data_directory = "path/to/clif/data",
  filetype = "parquet",
  timezone = "US/Central"
)
```

# Migrating CLIF 2.1 to 3.0

Sites still on CLIF 2.1 can upgrade their files to CLIF 3.0 with
`migrate_clif_2_1_to_3_0()`, which applies the bundled table crosswalks and writes the
migrated tables to an output directory. Use `dry_run = TRUE` to preview the changes
(including any timezone-handling concerns) without writing files.

```{r eval=FALSE}
migration_report <- migrate_clif_2_1_to_3_0(
  data_dir = "path/to/clif_2_1",
  output_dir = "path/to/clif_3_0",
  filetype = "parquet",
  dry_run = TRUE
)
```

Individual tables can be crosswalked in memory with `crosswalk_table_2_1_to_3_0()`
when you do not want to migrate a whole directory at once.

# Session info

```{r}
sessionInfo()
```
