---
title: "Getting Started with clifR"
author: "clifR Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with clifR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

# Introduction

**clifR** is an R port of the Python library
[clifpy](https://github.com/Common-Longitudinal-ICU-data-Format/clifpy). It
standardizes and analyses critical-care (ICU) data using the Common Longitudinal
ICU data Format (CLIF). clifR targets **CLIF 2.1** (the default) and can also load
and validate **CLIF 3.0** data, and it aims for behavioural parity with clifpy: the
same input produces the same results.

## What is CLIF?

The Common Longitudinal ICU data Format is a standardized relational schema designed to:

- facilitate multi-centre ICU research,
- enable reproducible critical-care analyses,
- provide consistent data validation, and
- support common clinical calculations (SOFA, Charlson, Elixhauser, and more).

CLIF defines a family of standardized tables covering patient demographics,
hospitalizations and transfers, vital signs and laboratory results, medications,
respiratory support, microbiology, procedures, and clinical assessments.

## Installation

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("AartikSarma/clifR")
```

```{r}
library(clifR)
library(dplyr)
library(ggplot2)
```

# The bundled demo dataset

clifR ships with a small, de-identified CLIF demo dataset so that every example in
these vignettes runs without any external files. It is the same demo data clifpy
bundles.

```{r}
list_demo_datasets()
```

Each table can be loaded on its own as a table object, or several tables can be
loaded together through the `ClifOrchestrator`.

# Loading a single table

The `load_demo_*()` helpers return CLIF table objects (R6 objects that inherit from
`BaseTable`). Under the hood these wrap the same classes you would construct with,
for example, `Patient$new(data_directory = ..., filetype = "parquet")` when working
with your own data.

```{r}
patient_table <- load_demo_patient()
patient_table
```

The parsed data lives in the `$df` field as a tibble:

```{r}
patient_table$df |>
  select(patient_id, sex_category, race_category, ethnicity_category) |>
  head()
```

Every table exposes `get_summary()`, which reports shape, memory use, numeric
summaries, and missing-data counts:

```{r}
patient_summary <- patient_table$get_summary()
patient_summary$num_rows
patient_summary$columns
```

Categorical columns can be tabulated by unique encounter with
`analyze_categorical_distributions()`:

```{r}
patient_distributions <- patient_table$analyze_categorical_distributions(save = FALSE)
patient_distributions$sex_category
```

# Validating a table

Validation checks the data against the CLIF schema for the configured version:
required columns, data types, permissible categorical values, datetime formats,
duplicates, and more. Call `validate()`, then `isvalid()` to check the outcome.

```{r}
hospitalization_table <- load_demo_hospitalization()
hospitalization_table$validate(verbose = TRUE)
hospitalization_table$isvalid()
```

When validation finds problems, the individual findings are stored in `$errors`:

```{r}
length(hospitalization_table$errors)
hospitalization_table$errors[[1]]
```

See the *Data quality and validation* vignette for the full validation engine and
report generation.

# Using the ClifOrchestrator

For multi-table work, `load_demo_clif()` returns a `ClifOrchestrator` with the
requested demo tables already loaded. (With your own data you would instead call
`ClifOrchestrator$new(data_directory = ..., filetype = ..., timezone = ..., clif_version = ...)`
followed by `initialize_tables()`.)

```{r}
clif <- load_demo_clif(tables = c("patient", "hospitalization", "adt", "vitals", "labs"))
clif
```

Loaded tables are available as fields, and `get_loaded_tables()` lists them:

```{r}
clif$get_loaded_tables()
nrow(clif$vitals$df)
```

Validate every loaded table at once:

```{r}
invisible(clif$validate_all(verbose = TRUE))
```

# A first exploration

## Hospital mortality and length of stay

The `Hospitalization` table provides convenience methods such as
`get_mortality_rate()` and `calculate_length_of_stay()`.

```{r}
hospitalization <- clif$hospitalization
sprintf("In-hospital mortality: %.1f%%", hospitalization$get_mortality_rate())

length_of_stay <- hospitalization$calculate_length_of_stay()
summary(length_of_stay$length_of_stay_days)
```

```{r fig.alt="Histogram of hospital length of stay in days"}
ggplot(length_of_stay, aes(x = length_of_stay_days)) +
  geom_histogram(bins = 30, fill = "#0072B2", colour = "white") +
  labs(
    title = "Hospital length-of-stay distribution",
    x = "Length of stay (days)",
    y = "Hospitalizations"
  ) +
  theme_minimal()
```

## Encounter stitching

Consecutive hospitalizations that are really one continuous encounter (a discharge
immediately followed by a transfer-driven readmission) can be linked with
`run_stitch_encounters()`. The mapping assigns each `hospitalization_id` to an
`encounter_block`.

```{r}
clif_with_adt <- load_demo_clif(tables = c("hospitalization", "adt"))
clif_with_adt$run_stitch_encounters()

encounter_mapping <- clif_with_adt$encounter_mapping
head(encounter_mapping)

sprintf(
  "%d hospitalizations collapse into %d encounter blocks",
  nrow(encounter_mapping),
  dplyr::n_distinct(encounter_mapping$encounter_block)
)
```

# Where to go next

- **Vitals and labs** — table-specific summary statistics, distributions, and ECDFs.
- **Clinical scores** — SOFA, Charlson (CCI), and Elixhauser.
- **Medication analysis** — dose-unit standardization for continuous infusions.
- **Wide datasets** — time-aligned wide format and hourly aggregation.
- **Microbiology** — culture/susceptibility tables and MDRO flags.
- **Data quality and validation** — the DQA engine, reports, and version migration.

# Additional resources

- CLIF specification: <https://clif-icu.com/data-dictionary>
- clifR on GitHub: <https://github.com/AartikSarma/clifR>
- Python clifpy: <https://github.com/Common-Longitudinal-ICU-data-Format/clifpy>

# Session info

```{r}
sessionInfo()
```
