---
title: "Medication Analysis"
author: "clifR Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Medication Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r}
library(clifR)
library(dplyr)
library(ggplot2)
```

# Overview

Continuous infusions in the `medication_admin_continuous` table arrive in a mix of
dose units: `mcg/kg/min`, `mg/hr`, `units/hr`, `mL/hr`, and so on. Before infusion
rates can be compared across patients (or fed into SOFA), they must be standardized
to a consistent preferred unit per medication category. clifR ports clifpy's DuckDB
unit-conversion pipeline for exactly this purpose.

The two entry points are:

- `convert_dose_units_by_med_category()` — the free function; operates on a medication
  data frame directly.
- `ClifOrchestrator$convert_dose_units_for_continuous_meds()` — the orchestrator
  wrapper; loads the medication and weight data it needs and calls the free function.

```{r}
clif <- load_demo_clif(tables = c("vitals", "medication_admin_continuous"))
medications <- clif$medication_admin_continuous

sort(unique(medications$df$med_category))
```

The units present in the demo data:

```{r}
medications$df |>
  count(med_dose_unit, sort = TRUE)
```

# Weight-based units need patient weight

Rate units such as `mcg/kg/min` are weight-normalized, so conversion needs a
`weight_kg` per hospitalization. When the medication table has no `weight_kg` column,
clifR derives the most recent weight from the vitals table automatically. That is why
we loaded `vitals` alongside the medication table above.

# Converting with the orchestrator

`convert_dose_units_for_continuous_meds()` takes a named list mapping each medication
category to its preferred unit. Two arguments matter for a clean run on real-world
data:

- `override = TRUE` — some categories (for example electrolyte drips measured in
  `mEq/min`) have no acceptable rate-unit target; `override` downgrades those from a
  hard error to a warning and leaves the affected rows unconverted.
- `save_to_table = FALSE` — returns the converted frame and conversion counts as a
  list, rather than writing them onto the (locked) table object.

```{r}
vasopressor_units <- list(
  norepinephrine = "mcg/kg/min",
  epinephrine    = "mcg/kg/min",
  phenylephrine  = "mcg/kg/min",
  dopamine       = "mcg/kg/min",
  dobutamine     = "mcg/kg/min",
  propofol       = "mcg/kg/min"
)

conversion <- clif$convert_dose_units_for_continuous_meds(
  preferred_units = vasopressor_units,
  override = TRUE,
  save_to_table = FALSE
)

names(conversion)
```

## Conversion status

The `counts` element tabulates how each original unit was handled. Categories with a
recognized rate unit convert successfully; the rest are reported as not recognized.

```{r}
conversion$counts |>
  count(`_convert_status`, wt = count, name = "n_rows") |>
  arrange(desc(n_rows))
```

## The converted doses

The converted frame keeps the original columns and adds the standardized dose and its
unit. For norepinephrine, every demo record was already in `mcg/kg/min`, so the values
pass through unchanged:

```{r}
conversion$converted |>
  filter(med_category == "norepinephrine") |>
  select(med_category, med_dose, med_dose_unit, med_dose_converted, med_dose_unit_converted) |>
  head()
```

```{r fig.alt="Distribution of standardized norepinephrine infusion rates"}
conversion$converted |>
  filter(med_category == "norepinephrine", !is.na(med_dose_converted)) |>
  ggplot(aes(x = med_dose_converted)) +
  geom_histogram(bins = 30, fill = "#D55E00", colour = "white") +
  labs(
    title = "Norepinephrine infusion rate",
    x = "Dose (mcg/kg/min)",
    y = "Administration records"
  ) +
  theme_minimal()
```

# Converting with the free function

`convert_dose_units_by_med_category()` is useful when you already hold the medication
and vitals data frames and want to control exactly which rows are processed. Here we
restrict to sedatives and provide the vitals frame explicitly for the weight lookup.

```{r}
sedative_doses <- medications$df |>
  filter(med_category %in% c("propofol", "dexmedetomidine", "fentanyl", "midazolam"))

sedative_conversion <- convert_dose_units_by_med_category(
  med_df = sedative_doses,
  vitals_df = clif$vitals$df,
  preferred_units = list(
    propofol        = "mcg/kg/min",
    dexmedetomidine = "mcg/kg/hr",
    fentanyl        = "mcg/hr",
    midazolam       = "mg/hr"
  ),
  override = TRUE
)

sedative_conversion$counts |>
  select(med_category, med_dose_unit, med_dose_unit_converted, `_convert_status`, count) |>
  arrange(med_category)
```

Acceptable target units follow clifpy's canonical short forms (for example `mcg/hr`,
`mg/hr`, `mcg/kg/min`, `u/hr`). Requesting a unit the pipeline cannot build — or a
category whose base unit has no valid rate form — is what triggers the "not
recognized" status and, without `override = TRUE`, an error.

# Session info

```{r}
sessionInfo()
```
