---
title: "Microbiology Data Analysis"
author: "clifR Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Microbiology Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This vignette demonstrates analysis of microbiological data including culture results, molecular testing, and antibiotic susceptibility patterns. These data are crucial for:

- Infection epidemiology
- Antibiotic stewardship
- Quality improvement initiatives
- Resistance tracking

# Setup

```{r eval=FALSE}
library(clifR)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)

# Initialize orchestrator
orchestrator <- ClifOrchestrator$new(
  data_directory = "path/to/clif/data",
  filetype = "csv",
  timezone = "America/New_York"
)

# Load microbiology tables
orchestrator$initialize_tables(
  tables = c(
    "patient", "hospitalization",
    "microbiology_culture",
    "microbiology_nonculture",
    "microbiology_susceptibility"
  ),
  validate = TRUE
)
```

# Culture-Based Microbiology

## Basic Culture Analysis

```{r eval=FALSE}
culture <- orchestrator$microbiology_culture

# View structure
head(culture$df)

# Summary
culture$summarize()

# Get distribution by specimen source
fluid_distribution <- culture$get_fluid_distribution()
print(fluid_distribution)
```

## Blood Cultures

```{r eval=FALSE}
# Get all blood culture results
blood_cultures <- culture$get_blood_cultures()

# Positive blood cultures
positive_blood_cultures <- blood_cultures %>%
  filter(organism_category != "no_growth")

# Positivity rate
blood_culture_positivity <- nrow(positive_blood_cultures) / nrow(blood_cultures) * 100
cat(sprintf("Blood culture positivity rate: %.1f%%\n", blood_culture_positivity))

# Most common blood culture organisms
blood_culture_organisms <- positive_blood_cultures %>%
  group_by(organism_category) %>%
  summarize(
    n_cultures = n(),
    n_patients = n_distinct(patient_id),
    .groups = "drop"
  ) %>%
  arrange(desc(n_cultures)) %>%
  head(10)

print(blood_culture_organisms)

# Visualize
ggplot(blood_culture_organisms, aes(x = reorder(organism_category, n_cultures), y = n_cultures)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Blood Culture Organisms",
    x = "Organism",
    y = "Number of Positive Cultures"
  ) +
  theme_minimal()
```

## Respiratory Cultures

```{r eval=FALSE}
# Get respiratory culture results
respiratory_cultures <- culture$get_respiratory_cultures()

# Common respiratory pathogens
resp_pathogens <- c(
  "pseudomonas_aeruginosa",
  "staphylococcus_aureus",
  "acinetobacter_baumannii",
  "klebsiella_pneumoniae",
  "haemophilus_influenzae"
)

respiratory_organisms <- respiratory_cultures %>%
  filter(organism_category %in% resp_pathogens) %>%
  group_by(organism_category) %>%
  summarize(
    n = n(),
    n_patients = n_distinct(patient_id),
    .groups = "drop"
  ) %>%
  arrange(desc(n))

print(respiratory_organisms)
```

## Organism Distribution

```{r eval=FALSE}
# Overall organism distribution (excluding no growth)
organism_dist <- culture$get_organism_distribution() %>%
  filter(organism_category != "no_growth") %>%
  head(20)

# Gram-positive vs Gram-negative
# (This would require organism group classification)
organism_groups <- culture$get_organism_distribution(by_group = TRUE)
print(organism_groups)
```

## Culture Turnaround Time

```{r eval=FALSE}
# Calculate turnaround times
tat_data <- culture$calculate_turnaround_time()

# Summary statistics
tat_summary <- tat_data %>%
  summarize(
    median_collection_to_result = median(collection_to_result_hours, na.rm = TRUE),
    median_order_to_result = median(order_to_result_hours, na.rm = TRUE),
    pct_result_within_48h = mean(order_to_result_hours <= 48, na.rm = TRUE) * 100
  )

print(tat_summary)

# Turnaround time by specimen type
tat_by_fluid <- tat_data %>%
  group_by(fluid_category) %>%
  summarize(
    n = n(),
    median_tat = median(collection_to_result_hours, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n >= 10) %>%  # Only fluids with sufficient samples
  arrange(desc(median_tat))

print(tat_by_fluid)
```

## Multi-Drug Resistant Organisms (MDROs)

Identify patients with MDR organisms (requires susceptibility data):

```{r eval=FALSE}
# Common MDROs
mdro_organisms <- c(
  "staphylococcus_aureus",  # May be MRSA
  "enterococcus_faecium",   # May be VRE
  "pseudomonas_aeruginosa", # May be MDR
  "acinetobacter_baumannii",# Often MDR
  "klebsiella_pneumoniae"   # May be ESBL or CRE
)

mdro_cultures <- culture$df %>%
  filter(organism_category %in% mdro_organisms)

cat(sprintf("%d cultures with potential MDROs\n", nrow(mdro_cultures)))
```

# Non-Culture Molecular Testing

## Basic Nonculture Analysis

```{r eval=FALSE}
nonculture <- orchestrator$microbiology_nonculture

# View structure
head(nonculture$df)

# Summary
nonculture$summarize()

# Test distribution
test_distribution <- nonculture$get_organism_distribution()
print(test_distribution)
```

## COVID-19 Testing

```{r eval=FALSE}
# Get all COVID-19 test results
covid_tests <- nonculture$get_covid_tests()

# Overall positivity
covid_positivity <- covid_tests %>%
  summarize(
    n_tests = n(),
    n_positive = sum(result_category == "detected", na.rm = TRUE),
    positivity_rate = n_positive / n_tests * 100
  )

print(covid_positivity)

# Positive tests over time
covid_positive_timeline <- covid_tests %>%
  filter(result_category == "detected") %>%
  mutate(
    week = floor_date(collect_dttm, "week")
  ) %>%
  group_by(week) %>%
  summarize(n_positive = n(), .groups = "drop")

ggplot(covid_positive_timeline, aes(x = week, y = n_positive)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 2) +
  labs(
    title = "COVID-19 Positive Tests Over Time",
    x = "Week",
    y = "Number of Positive Tests"
  ) +
  theme_minimal()
```

## C. difficile Testing

```{r eval=FALSE}
# Get C. diff test results
cdiff_tests <- nonculture$get_cdiff_tests()

# Positivity rate
cdiff_summary <- cdiff_tests %>%
  summarize(
    n_tests = n(),
    n_positive = sum(result_category == "detected", na.rm = TRUE),
    positivity_rate = n_positive / n_tests * 100,
    n_patients_tested = n_distinct(patient_id),
    n_patients_positive = n_distinct(patient_id[result_category == "detected"])
  )

print(cdiff_summary)

# Patients with recurrent C. diff (multiple positive tests)
recurrent_cdiff <- cdiff_tests %>%
  filter(result_category == "detected") %>%
  group_by(patient_id) %>%
  summarize(
    n_positive_tests = n(),
    .groups = "drop"
  ) %>%
  filter(n_positive_tests > 1)

cat(sprintf("%d patients with recurrent C. difficile\n", nrow(recurrent_cdiff)))
```

## Test Turnaround Time

```{r eval=FALSE}
# Calculate molecular test TAT
nonculture_tat <- nonculture$calculate_turnaround_time()

# Summary - molecular tests should be faster than cultures
nonculture_tat_summary <- nonculture_tat %>%
  summarize(
    median_collection_to_result = median(collection_to_result_hours, na.rm = TRUE),
    median_order_to_result = median(order_to_result_hours, na.rm = TRUE),
    pct_result_within_24h = mean(order_to_result_hours <= 24, na.rm = TRUE) * 100
  )

print(nonculture_tat_summary)
```

# Antibiotic Susceptibility

## Basic Susceptibility Analysis

```{r eval=FALSE}
susceptibility <- orchestrator$microbiology_susceptibility

# View structure
head(susceptibility$df)

# Summary
susceptibility$summarize()

# Antimicrobial testing distribution
antimicrobial_summary <- susceptibility$get_antimicrobial_summary() %>%
  head(20)

print(antimicrobial_summary)
```

## Antibiogram Creation

Hospital-level antibiogram (% susceptible for common bugs/drugs):

```{r eval=FALSE}
# Create antibiogram for common antibiotics
antibiogram <- susceptibility$get_antibiogram(common_only = TRUE)
print(antibiogram)

# Visualize antibiogram
ggplot(antibiogram, aes(x = reorder(antimicrobial_category, -pct_susceptible), y = pct_susceptible)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 90, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    title = "Hospital Antibiogram - % Susceptible",
    x = "Antibiotic",
    y = "% Susceptible"
  ) +
  theme_minimal()
```

## Organism-Specific Susceptibility

```{r eval=FALSE}
# Link susceptibility with culture data
susceptibility_with_organism <- susceptibility$join_with_cultures(
  culture_data = culture$df
)

# E. coli susceptibility patterns
ecoli_susceptibility <- susceptibility_with_organism %>%
  filter(organism_category == "escherichia_coli") %>%
  group_by(antimicrobial_category) %>%
  summarize(
    n_tested = n(),
    pct_susceptible = mean(susceptibility_category == "susceptible", na.rm = TRUE) * 100,
    pct_resistant = mean(susceptibility_category == "non_susceptible", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  filter(n_tested >= 10) %>%
  arrange(desc(pct_susceptible))

print(ecoli_susceptibility)
```

## Multi-Drug Resistant Organisms

```{r eval=FALSE}
# Identify MDR organisms (resistant to ≥3 antibiotic classes)
mdr_organisms <- susceptibility$get_mdr_organisms(min_resistances = 3)

# How many MDR organisms?
cat(sprintf("%d organism isolates with MDR patterns\n", nrow(mdr_organisms)))

# Most resistant organisms
top_mdr <- mdr_organisms %>%
  arrange(desc(n_resistances)) %>%
  head(10)

print(top_mdr)

# MDR organism types
if (nrow(mdr_organisms) > 0) {
  mdr_with_organism <- mdr_organisms %>%
    inner_join(
      culture$df %>% select(organism_id, organism_category, organism_group),
      by = "organism_id"
    )

  mdr_by_type <- mdr_with_organism %>%
    group_by(organism_category) %>%
    summarize(
      n_mdr_isolates = n(),
      mean_resistances = mean(n_resistances),
      .groups = "drop"
    ) %>%
    arrange(desc(n_mdr_isolates))

  print(mdr_by_type)
}
```

## Resistance Trends

```{r eval=FALSE}
# Track resistance over time for key pathogens
# Example: E. coli resistance to fluoroquinolones

resistance_trends <- susceptibility_with_organism %>%
  filter(
    organism_category == "escherichia_coli",
    antimicrobial_category == "ciprofloxacin"
  ) %>%
  mutate(
    month = floor_date(collect_dttm, "month")
  ) %>%
  group_by(month) %>%
  summarize(
    n_tested = n(),
    pct_resistant = mean(susceptibility_category == "non_susceptible", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  filter(n_tested >= 5)  # Only months with adequate sample size

ggplot(resistance_trends, aes(x = month, y = pct_resistant)) +
  geom_line(color = "red", size = 1) +
  geom_point(size = 2) +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  labs(
    title = "E. coli Ciprofloxacin Resistance Over Time",
    x = "Month",
    y = "% Resistant"
  ) +
  theme_minimal()
```

# Combined Analyses

## Infection Epidemiology

```{r eval=FALSE}
# Identify patients with positive cultures
infected_patients <- culture$df %>%
  filter(organism_category != "no_growth") %>%
  distinct(patient_id, hospitalization_id)

# Join with hospitalization outcomes
hosp_data <- orchestrator$hospitalization$df

infection_outcomes <- infected_patients %>%
  inner_join(
    hosp_data %>% select(hospitalization_id, hospital_expire_flag, length_of_stay_days),
    by = "hospitalization_id"
  )

# Compare to non-infected patients
infection_comparison <- hosp_data %>%
  left_join(infected_patients, by = "hospitalization_id") %>%
  mutate(has_positive_culture = !is.na(patient_id)) %>%
  group_by(has_positive_culture) %>%
  summarize(
    n = n(),
    mortality_rate = mean(hospital_expire_flag, na.rm = TRUE) * 100,
    median_los = median(length_of_stay_days, na.rm = TRUE),
    .groups = "drop"
  )

print(infection_comparison)
```

## Sepsis Identification

Combine blood culture results with clinical data:

```{r eval=FALSE}
# Patients with positive blood cultures (potential bacteremia/sepsis)
bacteremia_patients <- blood_cultures %>%
  filter(organism_category != "no_growth") %>%
  group_by(hospitalization_id) %>%
  summarize(
    first_positive_blood_culture = min(collect_dttm),
    organisms = paste(unique(organism_category), collapse = ", "),
    .groups = "drop"
  )

# Could join with:
# - Medication data (vasopressor timing)
# - Labs (lactate levels)
# - SOFA scores
# To identify sepsis cohort
```

## Antibiotic Appropriateness

Evaluate empiric vs. targeted therapy:

```{r eval=FALSE}
# This would require linking:
# 1. Culture collection time
# 2. Culture result time
# 3. Antibiotic administration times
# 4. Susceptibility results

# Example framework:
antibiotic_appropriateness <- culture$df %>%
  filter(organism_category != "no_growth") %>%
  select(hospitalization_id, organism_id, collect_dttm, result_dttm, organism_category)

# Join with susceptibility
appropriateness_data <- antibiotic_appropriateness %>%
  inner_join(
    susceptibility$df,
    by = "organism_id"
  )

# Join with antibiotic administration (from medication_admin_intermittent)
# Check if patient was on an appropriate antibiotic
```

# Quality Metrics

## Culture Collection Compliance

```{r eval=FALSE}
# Cultures per admission
cultures_per_admission <- culture$df %>%
  group_by(hospitalization_id) %>%
  summarize(
    n_cultures = n(),
    n_blood_cultures = sum(fluid_category == "blood_buffy"),
    .groups = "drop"
  )

# How many admissions had blood cultures drawn?
hosp_with_cultures <- hosp_data %>%
  left_join(cultures_per_admission, by = "hospitalization_id") %>%
  mutate(
    had_blood_culture = !is.na(n_blood_cultures) & n_blood_cultures > 0
  )

blood_culture_rate <- mean(hosp_with_cultures$had_blood_culture, na.rm = TRUE) * 100
cat(sprintf("%.1f%% of admissions had blood cultures\n", blood_culture_rate))
```

## Contamination Rates

Identify likely contaminants in blood cultures:

```{r eval=FALSE}
# Common contaminants
contaminants <- c(
  "staphylococcus_epidermidis",
  "staphylococcus_hominis",
  "corynebacterium_sp",
  "bacillus_sp",
  "cutibacterium_acnes"
)

# Contamination rate
blood_culture_contamination <- blood_cultures %>%
  filter(organism_category != "no_growth") %>%
  mutate(
    likely_contaminant = organism_category %in% contaminants
  ) %>%
  summarize(
    n_positive = n(),
    n_contaminants = sum(likely_contaminant),
    contamination_rate = n_contaminants / n_positive * 100
  )

print(blood_culture_contamination)
```

## Stewardship Metrics

```{r eval=FALSE}
# Days of therapy (DOT) per 1000 patient-days
# This would require:
# - Antibiotic administration data
# - Total patient-days from hospitalization

# Resistance rates for key organisms
key_organism_resistance <- susceptibility_with_organism %>%
  filter(organism_category %in% c(
    "escherichia_coli",
    "klebsiella_pneumoniae",
    "pseudomonas_aeruginosa",
    "staphylococcus_aureus"
  )) %>%
  group_by(organism_category) %>%
  summarize(
    n_tested = n(),
    resistance_rate = mean(susceptibility_category == "non_susceptible", na.rm = TRUE) * 100,
    .groups = "drop"
  )

print(key_organism_resistance)
```

# Summary

This vignette covered:

- Culture-based microbiology (blood cultures, respiratory cultures, organisms)
- Non-culture molecular testing (COVID-19, C. diff)
- Antibiotic susceptibility testing and antibiograms
- Multi-drug resistant organisms
- Infection epidemiology
- Quality and stewardship metrics

Microbiology data provides critical insights for infection control, antibiotic stewardship, and patient outcomes!

# Session Info

```{r eval=FALSE}
sessionInfo()
```
