---
title: "Microbiology Data Analysis"
author: "clifR Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Microbiology Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r}
library(clifR)
library(dplyr)
```

# Overview

CLIF represents microbiology in three related tables:

- `microbiology_culture` — organisms recovered from cultures, keyed by
  `organism_id`, `hospitalization_id`, and `organism_category`.
- `microbiology_susceptibility` — antimicrobial susceptibility results per
  `organism_id`, with `antimicrobial_category` and `susceptibility_category`.
- `microbiology_nonculture` — non-culture diagnostics (antigen/PCR assays).

clifR builds one derived analysis on top of these: **multidrug-resistant organism
(MDRO) flags**, via `calculate_mdro_flags()`.

> These microbiology tables are **not part of the bundled demo dataset**, so the
> table-loading snippets below are shown but not executed. The MDRO calculation is
> then demonstrated on a small illustrative input so you can see its real output.

# Loading microbiology tables

With your own CLIF data you would load the tables through the orchestrator (or
construct the table classes directly):

```{r eval=FALSE}
clif <- ClifOrchestrator$new(
  data_directory = "path/to/clif/data",
  filetype = "parquet",
  timezone = "US/Central",
  clif_version = "2.1"
)
clif$initialize_tables(tables = c(
  "microbiology_culture",
  "microbiology_susceptibility"
))

culture_table <- clif$microbiology_culture
susceptibility_table <- clif$microbiology_susceptibility

# Validate against the CLIF schema
culture_table$validate()
culture_table$isvalid()
```

The `MicrobiologyCulture` class adds helpers for exploring the organism vocabulary,
for example category-to-name mappings and the most frequent fluid/organism
combinations:

```{r eval=FALSE}
culture_table$organism_cat_name_map(include_counts = TRUE)
culture_table$top_fluid_org_outliers(level = "organism_category")
```

# MDRO flags

`calculate_mdro_flags()` merges culture and susceptibility results and applies the
resistance definitions in the bundled `mdro.yaml` configuration to flag
multidrug-resistant organisms. It works per organism; the configuration currently
defines *Pseudomonas aeruginosa*.

It accepts either table objects or plain data frames with the required columns. Here
we supply small illustrative frames describing one organism and its susceptibility
panel. In practice these come from the loaded microbiology tables.

```{r}
culture <- tibble::tibble(
  hospitalization_id = "H1",
  organism_id = "org1",
  organism_category = "pseudomonas_aeruginosa"
)

antimicrobial_panel <- c(
  "gentamicin", "tobramycin", "amikacin",       # aminoglycosides
  "meropenem",                                   # antipseudomonal carbapenems
  "ceftazidime", "cefepime",                     # antipseudomonal cephalosporins
  "ciprofloxacin", "levofloxacin",               # antipseudomonal fluoroquinolones
  "piperacillin_tazobactam",                     # penicillins + beta-lactamase inhib.
  "aztreonam",                                   # monobactams
  "fosfomycin",                                  # phosphonic acids
  "colistin"                                     # polymyxins
)

susceptibility <- tibble::tibble(
  organism_id = "org1",
  antimicrobial_category = antimicrobial_panel,
  susceptibility_category = c(
    "susceptible", "susceptible", "susceptible", # aminoglycosides
    "resistant",                                 # carbapenem
    "resistant", "resistant",                    # cephalosporins
    "resistant", "resistant",                    # fluoroquinolones
    "susceptible",                               # pip/tazo
    "susceptible",                               # aztreonam
    "susceptible",                               # fosfomycin
    "susceptible"                                # colistin
  )
)
```

```{r}
mdro_flags <- calculate_mdro_flags(
  culture = culture,
  susceptibility = susceptibility,
  organism_name = "pseudomonas_aeruginosa"
)

names(mdro_flags)
```

The result has one row per `(hospitalization_id, organism_id)`, with the per-agent
result, per antimicrobial-group resistance indicators, and the resistance-category
flags defined for the organism — `MDR`, `XDR`, `PDR`, and `DTR` for *P. aeruginosa*:

```{r}
mdro_flags |>
  select(hospitalization_id, organism_id, MDR, XDR, PDR, DTR)
```

```{r}
# Per-antimicrobial-group resistance for this isolate
mdro_flags |>
  select(hospitalization_id, organism_id, ends_with("_group"))
```

`calculate_mdro_flags()` warns when the panel is missing agents that a resistance
category depends on: an organism cannot be flagged for a category whose required
agents were never tested. Supplying a complete panel (as above) resolves those
warnings. A custom configuration can be provided through the `config_path` argument.

# Session info

```{r}
sessionInfo()
```
