---
title: "Working with Vitals and Labs"
author: "clifR Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Vitals and Labs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r}
library(clifR)
library(dplyr)
library(tidyr)
library(ggplot2)

# Okabe-Ito palette for discrete series (project convention).
okabe_ito <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
  "#0072B2", "#D55E00", "#CC79A7", "#000000"
)
```

# Overview

The `Vitals` and `Labs` tables hold the two highest-volume streams of ICU
measurements. Both inherit the generic `BaseTable` behaviour (validation, summaries,
categorical distributions, stratified ECDFs) and add a few table-specific helpers.
This vignette uses the bundled CLIF demo data throughout.

```{r}
vitals_table <- load_demo_vitals()
labs_table <- load_demo_labs()

vitals_table
labs_table
```

# Vital signs

## Which vitals are recorded?

```{r}
sort(unique(vitals_table$df$vital_category))
```

## Per-category summary statistics

`get_vital_summary_stats()` returns count, mean, standard deviation, min/max, and
quartiles for each `vital_category`.

```{r}
vital_summary <- vitals_table$get_vital_summary_stats()
vital_summary
```

## Filtering to one vital

`filter_by_vital_category()` returns the rows for a single vital as a tibble.

```{r}
heart_rate <- vitals_table$filter_by_vital_category("heart_rate")
nrow(heart_rate)
summary(heart_rate$vital_value)
```

## Distribution of a vital

```{r fig.alt="Boxplots of selected vital-sign distributions"}
selected_vitals <- c("heart_rate", "sbp", "dbp", "map", "respiratory_rate", "spo2")

vitals_table$df |>
  filter(vital_category %in% selected_vitals) |>
  ggplot(aes(x = vital_category, y = vital_value, fill = vital_category)) +
  geom_boxplot(outlier.alpha = 0.1) +
  scale_fill_manual(values = okabe_ito, guide = "none") +
  labs(
    title = "Distribution of selected vital signs",
    x = NULL,
    y = "Measured value"
  ) +
  theme_minimal()
```

# Laboratory results

## Per-category summary statistics

`get_lab_category_stats()` summarizes each `lab_category`, including a missing-value
percentage alongside the numeric summary.

```{r}
lab_stats <- labs_table$get_lab_category_stats()
lab_stats |>
  arrange(desc(count)) |>
  head(10)
```

## Specimen sources

```{r}
labs_table$get_lab_specimen_stats() |>
  head()
```

# Categorical distributions

`analyze_categorical_distributions()` counts distinct encounters per category value
for every categorical column defined in the schema; `plot_categorical_distributions()`
draws the matching bar plots (returning `ggplot` objects). Here we look at the lab
specimen category.

```{r}
lab_distributions <- labs_table$analyze_categorical_distributions(save = FALSE)
names(lab_distributions)
```

```{r fig.alt="Bar plot of lab specimen category distribution", fig.height=5}
specimen_plots <- labs_table$plot_categorical_distributions(
  columns = "lab_specimen_category",
  save = FALSE
)
specimen_plots[["lab_specimen_category"]]
```

# Stratified empirical CDFs

`calculate_stratified_ecdf()` computes, for each category value, the empirical
cumulative distribution of a numeric column: the sorted values paired with their
cumulative probability. This is a compact way to compare the full distribution of a
measurement across categories.

```{r}
vital_ecdf <- vitals_table$calculate_stratified_ecdf(
  value_column = "vital_value",
  category_column = "vital_category",
  category_values = c("heart_rate", "respiratory_rate", "spo2"),
  save = FALSE
)
names(vital_ecdf)
head(vital_ecdf$heart_rate)
```

```{r fig.alt="Empirical CDF curves for three vital signs"}
bind_rows(vital_ecdf) |>
  ggplot(aes(x = vital_value, y = cumulative_probability, colour = vital_category)) +
  geom_step() +
  scale_colour_manual(values = okabe_ito, name = "Vital") +
  labs(
    title = "Empirical CDFs by vital category",
    x = "Measured value",
    y = "Cumulative probability"
  ) +
  theme_minimal()
```

The same method works on `Labs`. Here we compare the distribution of a few
electrolytes:

```{r fig.alt="Empirical CDF curves for selected labs"}
lab_ecdf <- labs_table$calculate_stratified_ecdf(
  value_column = "lab_value_numeric",
  category_column = "lab_category",
  category_values = c("sodium", "potassium", "chloride"),
  save = FALSE
)

bind_rows(lab_ecdf) |>
  ggplot(aes(x = lab_value_numeric, y = cumulative_probability, colour = lab_category)) +
  geom_step() +
  scale_colour_manual(values = okabe_ito, name = "Lab") +
  labs(
    title = "Empirical CDFs by lab category",
    x = "Result value",
    y = "Cumulative probability"
  ) +
  theme_minimal()
```

# Session info

```{r}
sessionInfo()
```
