---
title: "Wide Datasets and Time-Series Analysis"
author: "clifR Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wide Datasets and Time-Series Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r}
library(clifR)
library(dplyr)
library(tidyr)
library(ggplot2)

okabe_ito <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
  "#0072B2", "#D55E00", "#CC79A7", "#000000"
)
```

# Overview

CLIF tables are stored in a narrow (long) form: one row per measurement, with a
category column naming the measurement and a value column holding it. Most analyses
instead want a **wide, time-aligned** view: one row per patient per timestamp, with
one column per measurement type. clifR builds this with `create_wide_dataset()` and
then aggregates it into fixed-width time windows with `convert_wide_to_hourly()`.

Both are DuckDB pipelines ported from clifpy for numerical parity.

# Building a wide dataset

`ClifOrchestrator$create_wide_dataset()` pivots the requested narrow tables and joins
them on a common event timeline. `category_filters` selects which category values to
keep from each pivot table (and becomes the resulting column names).

```{r}
clif <- load_demo_clif()

wide_dataset <- clif$create_wide_dataset(
  tables_to_load = c("vitals", "labs"),
  category_filters = list(
    vitals = c("heart_rate", "sbp", "map", "spo2"),
    labs   = c("creatinine", "sodium", "potassium")
  ),
  show_progress = FALSE
)

dim(wide_dataset)
names(wide_dataset)
```

Each measurement is now its own column, keyed by `hospitalization_id` and
`event_time`:

```{r}
wide_dataset |>
  select(hospitalization_id, event_time, heart_rate, map, spo2, creatinine) |>
  arrange(hospitalization_id, event_time) |>
  head(10)
```

Because measurements are recorded at different times, any single row is typically
sparse (a vitals row has no lab values, and vice versa). The next step fills that in
by aggregating within time windows.

# Aggregating into time windows

`convert_wide_to_hourly()` collapses the wide dataset into fixed-width windows per
patient. The `aggregation_config` maps an aggregation method to the columns it applies
to. Available methods include `mean`, `max`, `min`, `median`, `first`, `last`,
`boolean`, and `one_hot_encode`. `hourly_window` sets the window width in hours.

```{r}
hourly_dataset <- clif$convert_wide_to_hourly(
  aggregation_config = list(
    mean = c("heart_rate", "sbp", "map", "spo2"),
    max  = c("creatinine", "potassium"),
    min  = c("sodium")
  ),
  hourly_window = 6
)

dim(hourly_dataset)
names(hourly_dataset)
```

Aggregated columns are suffixed with their method, and each window carries its start
and end timestamps:

```{r}
hourly_dataset |>
  select(hospitalization_id, window_number, window_start_dttm,
         heart_rate_mean, map_mean, spo2_mean, creatinine_max) |>
  arrange(hospitalization_id, window_number) |>
  head(10)
```

# A worked example: one patient's trajectory

```{r}
example_hospitalization <- hourly_dataset |>
  count(hospitalization_id, sort = TRUE) |>
  slice(1) |>
  pull(hospitalization_id)

patient_windows <- hourly_dataset |>
  filter(hospitalization_id == example_hospitalization) |>
  arrange(window_number)

nrow(patient_windows)
```

```{r fig.alt="Windowed mean vital signs over time for one hospitalization"}
patient_windows |>
  select(window_start_dttm, heart_rate_mean, map_mean, spo2_mean) |>
  pivot_longer(-window_start_dttm, names_to = "measurement", values_to = "value") |>
  filter(!is.na(value)) |>
  ggplot(aes(x = window_start_dttm, y = value, colour = measurement)) +
  geom_line() +
  geom_point(size = 1) +
  scale_colour_manual(values = okabe_ito, name = NULL) +
  labs(
    title = sprintf("6-hour windowed vitals, hospitalization %s", example_hospitalization),
    x = "Window start",
    y = "Mean value"
  ) +
  theme_minimal()
```

# Notes

- `create_wide_dataset()` also stores its result on the orchestrator as `clif$wide_df`,
  so `convert_wide_to_hourly()` can be called without re-passing it.
- For very large datasets, `create_wide_dataset()` accepts `batch_size`,
  `memory_limit`, and `threads` to tune the DuckDB engine, and can write directly to
  disk via `output_format` / `save_to_data_location` instead of returning a data frame.
- `hourly_window = 1` gives true hourly windows; larger values give 4-hourly, 6-hourly,
  daily, etc.

# Session info

```{r}
sessionInfo()
```
